Exploring the tree of numerical semigroups

### 2014-01-06 Papier/article.tex
Modification of proof of Proposition 3.5 and small correction in sections <=3

A \emph{numerical semigroup} $S$ is a subset of $\NN$ containing $0$, close under addition and of finite complement in $\NN$.  
For example the set 
\begin{equation}
\label{E:NSG}
S_E=\{0,3,6,7,9,10\}\cup[12,+\infty[
\end{equation}
is a numerical semigroup.
The \emph{genus} of a numerical semigroup $S$, denoted by~$g(S)$, is the cardinality of $\NN\setminus S$, \ie, $g(S)=\card(\NN\setminus S)$. 
 For example the genus of $S_E$ defined in \eqref{E:NSG} is $6$,  the cardinality of $\{1,2,4,5,8,11\}$

For a given positive integer $g$, the number of numerical semigroups  is finite and is denoted by $n_g$. 
In  J.A. Sloane's \emph{on-line encyclopedia of integer}~\cite{OEIS} we find the values of $n_g$ for $g\leq 52$. 
These values have been obtain by M. Bras-Amor\'os (view \cite{BrasAmoros2008} for more details for $g\leq 50)$. 
On his home page~\cite{Delgado}, M. Delgado  gives the value of $n_{55}$ without specifying the values of $n_{53}$ and~$n_{54}$.

M.~Bras-Amor\'os used a depth first search exploration of the tree of numerical semigroups up to a given genus.
This tree was introduced by J.C. Rosales and al. in \cite{Rosales} and it is the subject of the Section~\ref{S:Tree}.
Starting with all the numerical semigroups of genus $49$ she obtained the number of numerical semigroups of genus $50$ in $18$ days on a pentium D runing at $3$GHz. 
In the package~\tt{NumericalSgs} \cite{NumericalSgps} of \tt{GAP} \cite{GAP}, M. Delgado together with P.A.~Garcia-Sanchez and J. Morais used the same method of exploration.

@@

\begin{defi}
A subset $X=\{x_1<x_2<...<x_n\}$ of a semigroup is a \emph{generating set} of $S$ if every element of $S$ can be express as a sum of elements in $X$. 
In this case we write $S=\left<x_1,...,x_\ell\right>$.
\end{defi}

@@

\begin{lem}[Lemma 2.3, Chap. I, \cite{BookNS}]
Let $S$ be a numerical semigroup. The set $\Irr(S)$ is the minimal generating set of $S$.
\end{lem}

@@

\begin{proof}
$i)$ Let $x$ ba a positive integer satisfying $x \geq c(S)+m(S)$. 
As the relations $x\geq c(S)$ and $x-m(S)$ hold, the integers $x$ and $x-m(S)$ belong to $S$.
Since $x=(x-m(S))+m(S)$ hold with $x-m(S)\not=0$ and $m(S)\not=0$, the integer $x$ could not be irreducible.

@@

$iii)$ Let $x$ be an element of $S\cap[0,c(S)-1]$. 
The integer $y=c(S)-1-x$ lies in $[0,c(S)-1]$. 
Moreover $y$ is not in $S$, for otherwise  we write $c(S)-1=y+x$ where $x,y$ are elements of $S$ implying $c(S)-1\in S$. 
In contradiction with the definition of conductor. 
Thus we define an involution 
\[
\begin{array}{rcl}
\psi:[0,c(S)-1]&\to&[0,c(S)-1]\\
x&\mapsto&c(S)-1-x
\end{array}
\]
that send $S\cap[0,c(S)-1]$ into $[0,c(S)-1]\setminus S$.
The cardinality of $[0,c(S)[\setminus S$ is exactly $g(S)$. Let us denote by $k$ the cardinality  of $S\cap[0,c(S)-1]$. 
Since $\psi$ is injective we must have $k\leq g(S)$. 
From the relation 
\[
[0,c(S)[=S\cap[0,c(S)[\sqcup[0,c(S)[\setminus S
\]
 we obtain $c(S)=k+g(S)$ and so $c(S)\leq 2 g(S)$.
\end{proof}

A consequence of Proposition~ \ref{P:Res}~$i)$ is that $\Irr(S)$ is finite of cardinality at most $c(S)+m(S)-1$.
In fact, using Apéry set, we can prove that  the cardinality of $\Irr(S)$ is at most $m(S)$. See Section 2 of Chapter I of \cite{BookNS} for more details.

Let  $S$ be a numerical semigroup. The set $T=S\cup\{c(S)-1\}$ is also a numerical  semigroup and its genus is $g(S)-1$. 
As  $[c(S)-1,+\infty[$ is included in $T$ we have $c(T)\leq c(S)-1$. 
Therefore every semigroup $S$ of genus $g$ can be obtained from a semigroup $T$ of genus $g-1$ by removing an element of~$T$ greater than or equal to $c(T)$.

@@

We construct the tree of numerical semigroups, denoted by $\mathcal{T}$ as follows. 
The root of the tree is the unique semigroup of genus $0$, \ie, $\left<1\right>$ that is equal to $\NN$. 
If~$S$ is a semigroup in the tree,  the sons of $S$ are exactly the semigroup~$S^x$ where $x$ belongs to $\Irr(S)\cap[c(s),+\infty[$. 
By convention, when depicting the tree, the numerical semigroup $S^x$ is in the left of $S^y$ if $x$ is smaller than~$y$.

The above remarks imply that a semigroup $S$ has depth $g$ in $\mathcal{T}$ if and only if its genus is~$g$, see Figure~\ref{F:Tree}.
We denote by $\mathcal{T}^{\leq g}$ the subtree of $\mathcal{T}$ resticted to all semigroup of genus $\leq g$.

@@

\[
\begin{tikzpicture}[level distance=2cm, inner sep=2mm,
  level/.style={sibling distance=1.55cm},
%  edge from parent/.style={draw}
  ]
    \node {$\sgnode{1}$}
    child {node {$\sgnode{2,3}$}
      child [sibling distance=4.5cm] {node {$\sgnode{3,4,5}$}
        child {node {$\sgnode{4,5,6,7}$}
          child [sibling distance=2cm]{node {$\sgnode{5,6,7,8,9}$}  edge from parent node [left] {4}}
          child [sibling distance=1.9cm]{node {$\sgnode{\gr 4,6,7,9}$}  edge from parent node [left] {5}}
          child {node {$\sgnode{\gr 4,\gr 5,7}$}  edge from parent node [right] {6}}
          child {node {$\sgnode{\gr 4,\gr 5,\gr 6}$}  edge from parent node [right] {7}}
          edge from parent node [left] {3}
        }
        child{{} edge from parent[draw=none]}
        child{{} edge from parent[draw=none]}
        child {node {$\sgnode{\gr 3,5,7}$}
          child {node {$\sgnode{\gr 3,7,8}$} edge from parent node [left] {5}}
          child {node {$\sgnode{\gr 3,\gr 5}$} edge from parent node [right] {7}}
          edge from parent node [left] {4}
        }
        child {node {$\sgnode{\gr 3,\gr 4}$}
          edge from parent node [right] {5}
        }
        edge from parent node [left] {2}
      }
      child [sibling distance=4.5cm] {node {$\sgnode{\gr2,5}$}
        child {node {$\sgnode{\gr2,7}$}
          child {node {$\sgnode{\gr2,9}$}
            edge from parent node [right] {7}
          }
          edge from parent node [right] {5}
        }
        edge from parent node [right] {3}
      }
      edge from parent node [left] {1}
    }
    ;
  \end{tikzpicture}
  \]
\caption{The first five layers of the tree $\mathcal{T}$ of numerical semigroups, corresponding to $\mathcal{T}^{\leq 5}$. A generator of a semigroup is in gray if it is not in $[c(S),+\infty[$. An edge between a semigroup $S$ and its son $S'$ is labelled by  $x$ is $S'$ is obtained from $S$ by removing~$x$, that is $S'=S^x$.}
\label{F:Tree}
\end{figure}

As the reader can check, the main difficulty to characterize the son of a semigroup is to determine its irreducible elements.
 In \cite{NumericalSgps}, the semigroup are given by their Ap\'ery set and then the main difficulty is to describe the Ap\'ery set of $S_x$  from this of $S$. 
 This approach is elegant but not sufficiently \emph{basic} for our optimisations.

@@

The aim of this section is to describe a new representation of numerical semigroups.
which is well suited to a efficient exploration of the tree~$\mathcal{T}$ of numerical semigroups.

@@

Let us show $ii)$. 
Assume $x$ is irreducible in $S$.
There cannot exist $y$ and $z$ in~$(S\setminus\{0\})^2$ such that $x=y+z$. 
The only possible decomposition of $x$ as a sum of two elements of $S$ is $x=0+x$. 
Hence, the set $D_S(x)$ is equal to~$\{0\}$ and we have $d_S(x)=1$. 
Conversely, let $x$ such that $d_S(x)=1$. 
By  $i)$ the  integer $x$ must be in $S$. 
As $x=0+x$ is always a decomposition of $x$ as a sum of two elements in $S$, we obtain $D_S(x)=\{0\}$.
If there exist $y$ and $z$ in $S$ such that $y\leq z$ and $x=y+z$ hold then $y$ lies in $D_S(x)$. 
This implies $y=0$ and $z=x$. 
Hence $x$ is irreducible in~$S$.
\end{proof}

@@

We now explain how to compute the $S$-decomposition numbers function of a numerical semigroup  from these of its father.

@@

\begin{proof}
Let $y$ be in $\NN\setminus\{0\}$. We have 
\[
D_{S^x}(y)=\{z \in S^x\ |\ y-z\in S^x\ \text{and}\  2z\leq y\}
\]
and $D_{S^x}(y)$ is a subset of $D_S(y)$.
We have $D_S(y)\setminus D_{S^x}(y)=E\sqcup F$ where
\begin{align*}
E&=\begin{cases}
\{x\} &\text{if $y-x\in S^x$ and $2x\leq y$,}\\
\emptyset & \text{otherwise.}
\end{cases}\\
F&=\{z\in S\ |\ y-z=x\ \text{and}\ 2z\leq y\}
\end{align*}
Since  the relation $y-x\in S^x$ can be rewritten as the conjunction of $y-x\in S$ and $y\not=2x$, we obtain
\[
E=\begin{cases}
\{x\} &\text{if $y-x\in S$ and $2x< y$,}\\
\emptyset & \text{otherwise.}
\end{cases}
\]
By Proposition~\ref{P:Dandd}, the relation $y-x\in S$ is equivalent to $y\geq x$ and $d_S(y-x)>0$, we have
\[
E=\begin{cases}
\{x\} &\text{if $y\geq x$ and $d_S(y-x)>0$ and $2x< y$,}\\
\emptyset & \text{otherwise.}
\end{cases}
\]
On the other hand, we have 
\begin{align*}
F&=\{z\in S\ |\ y-z=x\ \text{and}\ 2z\leq y\}\\
&=\{z\in S\ |\ z=y-x\ \text{and}\ y\leq 2x\}\\
&=\begin{cases}
\{y-x\} & \text{if $y-x\in S$ and $y\leq 2x$,}\\
\emptyset & \text{otherwise.}
\end{cases}
\end{align*}
As in the case of $E$, we get 
\[
F=\begin{cases}
\{y-x\} & \text{if $y\geq x$ and $d_S(y-x)>0$ and $y\leq 2x$,}\\
\emptyset & \text{otherwise.}
\end{cases}
\]
In $E$ we have the constraint $2x<y$ while in $F$ we have $y\leq 2x$, hence only one of the sets $E$ or $F$ can be non-empty. 
This implies
\[
D_S(y)\setminus D_{S^x}(y)=
\begin{cases}
\{x\} & \text{if $y\geq x$ and $d_S(y-x)>0$ and $y\leq 2x$,}\\
\{y-x\} & \text{if $y\geq x$ and $d_S(y-x)>0$ and $y>2x$,}\\
\emptyset & \text{otherwise.}
\end{cases}
\]
Therefore we obtain
\[
d_{S^x}(y)=\begin{cases}
d_S(y)-1&\text{if $y\geq x$ and $d_S(y-x)>0$,}\\
d_S(y)& \text{otherwise.}
\end{cases}
\]
\end{proof}

>>>>

A \emph{numerical semigroup} $S$ is a subset of $\NN$ containing $0$, close under addition and of finite complement in $\NN$.  
For example the set 
\begin{equation}
\label{E:NSG}
S_E=\{0,3,6,7,9,10\}\cup[12,+\infty[
\end{equation}
is a numerical semigroup.
The \emph{genus} of a numerical semigroup $S$, denoted by~$g(S)$, is the cardinality of $\NN\setminus S$.
 For example the genus of $S_E$ is $6$,  the cardinality of $\{1,2,4,5,8,11\}$

For a given positive integer $g$, the number of numerical semigroups of genus $g$ is finite and is denoted by $n_g$. 
In  J.A. Sloane's \emph{on-line encyclopedia of integer}~\cite{OEIS} we find the values of $n_g$ for $g\leq 52$. 
These values have been obtain by M. Bras-Amor\'os (view \cite{BrasAmoros2008} for more details for $g\leq 50)$. 
On his home page~\cite{Delgado}, M. Delgado  gives the value of $n_{55}$ without specifying the values of $n_{53}$ and~$n_{54}$.

M.~Bras-Amor\'os used a depth first search exploration of the tree of numerical semigroups up to a given genus.
This tree was introduced by J.C.~Rosales and al. in \cite{Rosales} and it is the subject of the Section~\ref{S:Tree}.
Starting with all the numerical semigroups of genus $49$ she obtained the number of numerical semigroups of genus $50$ in $18$ days on a pentium D runing at $3$GHz. 
In the package~\tt{NumericalSgs} \cite{NumericalSgps} of \tt{GAP} \cite{GAP}, M. Delgado together with P.A.~Garcia-Sanchez and J. Morais used the same method of exploration.

@@

\begin{defi}
A subset $X=\{x_1<x_2<...<x_n\}$ of a semigroup is a \emph{generating set} of $S$ if every element of $S$ can be express as a sum of elements in $X$. 
In this case we write $S=\left<x_1,...,x_n\right>$.
\end{defi}

@@

\begin{lem}[Lemma 2.3, Chap. I, \cite{BookNS}]
Let $S$ be a numerical semigroup. The set $\Irr(S)$ is the minimal generating set of $S$ relatively to the inclusion ordering.
\end{lem}

@@

\begin{proof}
$i)$ Let $x$ ba a positive integer satisfying $x \geq c(S)+m(S)$. 
As the integers $x$ and $x-m(S)$ are greater than $c(S)$, they belong to $S$.
Since $x$ equals $(x-m(S))+m(S)$ with $x-m(S)\not=0$ and $m(S)\not=0$, it could not be irreducible.

@@

$iii)$ Let $x$ be an element of $S\cap[0,c(S)-1]$. 
The integer $y=c(S)-1-x$ lies in $[0,c(S)-1]$. 
Moreover $y$ is not in $S$, for otherwise  we write $c(S)-1=y+x$ where $x,y$ are elements of $S$ implying $c(S)-1\in S$. 
In contradiction with the definition of conductor. 
Thus we define an involution 
\[
\begin{array}{rcl}
\psi:[0,c(S)-1]&\to&[0,c(S)-1]\\
x&\mapsto&c(S)-1-x
\end{array}
\]
that send $S\cap[0,c(S)-1]$ into $[0,c(S)-1]\setminus S$ which is  of cardinality $g(S)$. 
Let us denote by $k$ the cardinality  of $S\cap[0,c(S)-1]$. 
Since $\psi$ is injective we must have $k\leq g(S)$. 
From the relation 
\[
[0,c(S)[=S\cap[0,c(S)[\sqcup[0,c(S)[\setminus S
\]
 we obtain $c(S)=k+g(S)$ and so $c(S)\leq 2 g(S)$.
\end{proof}

A consequence of Proposition~ \ref{P:Res}~$i)$ is that $\Irr(S)$ is finite ant its cardinality is at most $c(S)+m(S)-1$.
More precisely, using Ap\'ery set, we can prove that  the cardinality of $\Irr(S)$ is at most $m(S)$. See Section 2 of Chapter I of~\cite{BookNS} for more details.

We now explain the construction of the tree of numerical semigroups.
Let~$S$ be a numerical semigroup. The set $T=S\cup\{c(S)-1\}$ is also a numerical  semigroup and its genus is $g(S)-1$. 
As  $[c(S)-1,+\infty[$ is included in $T$ we have $c(T)\leq c(S)-1$. 
Therefore every semigroup $S$ of genus $g$ can be obtained from a semigroup $T$ of genus $g-1$ by removing an element of~$T$ greater than or equal to $c(T)$.

@@

We construct the tree of numerical semigroups, denoted by $\mathcal{T}$ as follows. 
The root of the tree is the unique semigroup of genus $0$, \ie, $\left<1\right>$ that is equal to $\NN$. 
If~$S$ is a semigroup in the tree,  the sons of $S$ are exactly the semigroup~$S^x$ where $x$ belongs to $\Irr(S)\cap[c(s),+\infty[$. 
By convention, when depicting the tree, the numerical semigroup $S^x$ is in the left of $S^y$ if $x$ is smaller than~$y$. 
With this construction, a semigroup $S$ has depth $g$ in $\mathcal{T}$ if and only if its genus is~$g$, see Figure~\ref{F:Tree}.
We denote by $\mathcal{T}_{\leq g}$ the subtree of $\mathcal{T}$ resticted to all semigroup of genus $\leq g$.

@@

\[
\begin{tikzpicture}[level distance=2cm, inner sep=2mm,
  level/.style={sibling distance=1.55cm},
%  edge from parent/.style={draw}
  ]
    \node {$\sgnode{1}$}
    child {node {$\sgnode{2,3}$}
      child [sibling distance=4.5cm] {node {$\sgnode{3,4,5}$}
        child {node {$\sgnode{4,5,6,7}$}
          child [sibling distance=2cm]{node {$\sgnode{5,6,7,8,9}$}  edge from parent node [left] {4}}
          child [sibling distance=1.9cm]{node {$\sgnode{\gr 4,6,7,9}$}  edge from parent node [left] {5}}
          child {node {$\sgnode{\gr 4,\gr 5,7}$}  edge from parent node [right] {6}}
          child {node {$\sgnode{\gr 4,\gr 5,\gr 6}$}  edge from parent node [right] {7}}
          edge from parent node [left] {3}
        }
        child{{} edge from parent[draw=none]}
        child{{} edge from parent[draw=none]}
        child {node {$\sgnode{\gr 3,5,7}$}
          child {node {$\sgnode{\gr 3,7,8}$} edge from parent node [left] {5}}
          child {node {$\sgnode{\gr 3,\gr 5}$} edge from parent node [right] {7}}
          edge from parent node [left] {4}
        }
        child {node {$\sgnode{\gr 3,\gr 4}$}
          edge from parent node [right] {5}
        }
        edge from parent node [left] {2}
      }
      child [sibling distance=4.5cm] {node {$\sgnode{\gr2,5}$}
        child {node {$\sgnode{\gr2,7}$}
          child {node {$\sgnode{\gr2,9}$}
            edge from parent node [right] {7}
          }
          edge from parent node [right] {5}
        }
        edge from parent node [right] {3}
      }
      edge from parent node [left] {1}
    }
    ;
  \end{tikzpicture}
  \]
\caption{The first four layers of the tree $\mathcal{T}$ of numerical semigroups, corresponding to $\mathcal{T}_{\leq 4}$. A generator of a semigroup is in gray if it is not in $[c(S),+\infty[$. An edge between a semigroup $S$ and its son $S'$ is labelled by  $x$ is $S'$ is obtained from $S$ by removing~$x$, that is $S'=S^x$.}
\label{F:Tree}
\end{figure}

As the reader can check, the main difficulty to characterize the son of a semigroup is to determine its irreducible elements.
 In \cite{NumericalSgps}, the semigroup are given by their Ap\'ery set and then the difficulty is to describe the Ap\'ery set of $S_x$  from this of $S$. 
 This approach is elegant but not sufficiently \emph{basic} for our optimisations.

@@

The aim of this section is to describe a new representation of numerical semigroups.
which is well suited to an efficient exploration of the tree~$\mathcal{T}$ of numerical semigroups.

@@

Let us show $ii)$. 
Assume $x$ is irreducible in $S$.
There cannot exist $y$ and $z$ in~$(S\setminus\{0\})^2$ such that $x=y+z$. 
The only possible decomposition of $x$ as a sum of two elements of $S$ is $x=0+x$. 
Hence, the set $D_S(x)$ is equal to~$\{0\}$ and we have $d_S(x)=1$. 
Conversely, let $x$ such that $d_S(x)$ is $1$. 
By  $i)$ the  integer $x$ must be in $S$. 
As $x=0+x$ is always a decomposition of $x$ as a sum of two elements in $S$, we obtain $D_S(x)=\{0\}$.
If there exist $y$ and $z$ in $S$ such that $y\leq z$ and $x=y+z$ hold then $y$ lies in $D_S(x)$. 
This implies $y=0$ and $z=x$. 
Hence $x$ is irreducible in~$S$.
\end{proof}

@@

We now explain how to compute the $S$-decomposition numbers function of a numerical semigroup  from this of its father.

@@

\begin{proof}
Let $y$ be in $\NN\setminus\{0\}$. We have 
\[
D_{S^x}(y)=\{z \in S^x\ |\ y-z\in S^x\ \text{and}\  2z\leq y\}.
\]
We construct three sets as follow:
\begin{align*}
E&=\{z\in S^x\ |\ y-z=x\ \text{and}\ 2z\leq y\}\\
F&=\begin{cases}
\{x\}&\text{if $y-x\in S$ and $2x\leq y$},\\
\emptyset&\text{otherwise}.
\end{cases}
\end{align*}
We have the relation $D_{S}(y)=D_{S^x}(y)\sqcup E\sqcup F$ and so $d_S(y)$ is equal to $d_{S^x}(y)+\card(E)+\card(F)$.
and $D_{S^x}(y)$ is a subset of $D_S(y)$. We start by determining $E$. 
The relation $y-z=x$ implies $z=y-x$ and $2z\leq y$ becomes $y\leq 2x$. Thus we obtain
\[
E=\begin{cases}
\{y-x\}&\text{if $y-x\in S^x$ and $y\leq 2x$}\\
\emptyset&\text{otherwise}
\end{cases}
\] 
The relation $y-x\in S^x$ together with $y\leq2x$ is equivalent to $y-x\in S$ together with $y<2x$.
Therefore we obtain
\[
E=\begin{cases}
\{y-x\}&\text{if $y-x\in S$ and $y< 2x$}\\
\emptyset&\text{otherwise}
\end{cases}
\] 
Assume $y-x$ lies in $S$.  If $y<2x$ holds, we have $E=\{y-x\}$ and $F=\emptyset$. Otherwise, we obtain $E=\emptyset$ and $F=\{x\}$. 
Then if $y-x\in S$ holds, we have $\card(E)+\card(F)=1$ and so $d_{S}(y)$ is equal to $d_{S^x}(y)+1$.
If $y-x$ not belong to $S$, then $E$ and $F$ are empty implying $d_{S}(y)=d_{S^x}(y)$. As the condition $y-x\in S$ is equivalent to $y\geq x$ together with $d_S(y-x)>0$ by Proposition~\ref{P:Dandd} the result holds.
\end{proof}
